---
title: "Class 8: Halloween Candy Mini Project"
author: "Clarissa Savko (PID: A69028482)"
format: pdf
---

First I have to get my data into R!
```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)

```
Q1. How many different candy types are in this dataset?

```{r}
dim(candy)
```
There are 85 candy types.
Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
```{r}
candy[as.logical(candy$fruity),]
```

There are 38 fruity candies in the dataset.

Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Reese's Peanut Butter cup", ]$winpercent
```
My favorite candy is Reeses peanut butter cups! It's `win.percent()` value is 84.18029.

Q4. What is the winpercent value for “Kit Kat”?
76.7686
```{r}
candy["Kit Kat", ]$winpercent
candy["Tootsie Roll Snack Bars", ]$winpercent
```

Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
49.6535

```{r}
skimr::skim(candy)
```

Double colon is a shortcut for using a package to avoid using the library function.

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent has a way higher mean and SD than the other columns.

Q7. What do you think a zero and one represent for the candy$chocolate column?

0 would be if the candy does not contain chocolate (false). 1 would be the candy does contain chocolate (true).

Q8. Plot a histogram of winpercent values
```{r}
hist(candy$winpercent)
```


```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent)+
  geom_histogram()
```

Q9. Is the distribution of winpercent values symmetrical?
No, it has a tail to the right.

Q10. Is the center of the distribution above or below 50%?
The center is below 50%.

Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}
choc.ind <- as.logical(candy$chocolate)
fruit.ind <- as.logical(candy$fruity)
choc.win <- candy[choc.ind,]$winpercent
fruit.win <- candy[fruit.ind,]$winpercent
mean(choc.win)
mean(fruit.win)
```
choc.win is higher ranked than fruit candy.

Q12. Is this difference statistically significant?
```{r}
t.test(choc.win, fruit.win)
```
Yes! The p value is 2.871e-08.

Q13. What are the five least liked candy types in this set?
```{r}
x <- c(5, 2, 3, 6)
sort(x)
sort(x, decreasing = T)
```
```{r}
x
order(x)
x[order(x)]
```
```{r}
y <- c("D", "A", "E")
order(y)
```
```{r}
y[order(y)]
```
```{r}
inds <- order(candy$winpercent)
candy[inds,]
head(candy[inds,])
tail(candy[inds,])
```

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```
The 5 least liked candy types are Nik l Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.
Q14. What are the top 5 all time favorite candy types out of this set?
The top 5 candies are Reese's peanut butter cups, Reese's miniatures, Twix, Kit Kat, and Snickers.

Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
ggplot(candy)+
  aes(winpercent, rownames(candy))+
  geom_col()
```

```{r}
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy),winpercent))+
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
Now, for the first time, using this plot we can answer questions like:
- Q17. What is the worst ranked chocolate candy?
The worst ranked chocolate candy is Sixlets.

- Q18. What is the best ranked fruity candy?
Starburst is the highest ranked fruity candy.

```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniature seems to be the highest ranked for the least money.
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candies are Nik L Nip, Nestle Smarties, Ring pop, Hershey;s Krackel, and Hershey's milk chocolate. Nik L Nip is the least popular (and most expensive!) with a winpercent of 22.45.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Chocolate and fruity are anti-correlated.

Q23. Similarly, what two variables are most positively correlated?
Chocolate and winpercent are the most positively correlated.

```{r}
pca <- prcomp(candy, scale. = TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

#p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  #theme(legend.position = "none") +
  #labs(title="Halloween Candy PCA Space",
       #subtitle="Colored by type: chocolate bar (dark brown), chocolate other #(light brown), fruity (red), other (black)",
       #caption="Data from 538")
```
```{r}
#library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Candies that are fruity, hard, and pluribus are picked up strongly in PC1 in the positive direction. This makes sense! Skittles and starbust are hard, fruity, and come with multiple in one package.