---
title: "Class 5: Data Visualization with ggplot"
author: "Clarissa (PID: A69028482)"
format: pdf
---

## 
Use the function "install.packages()" to install ggplot2
Before I use any package, I have to load them with "library()" call like so:

```{r}
#install.packages("ggplot2")
library(ggplot2)
#ggplot(cars)
```

##
Use "head()" as good practice to only render the first 6 rows
```{r}
head(cars)
```
##
There is always the "base R" graphics system, i.e. "plot()"
```{r}
plot(cars)
```
## To use ggplot I need to spell out at least 3 things:
-data (stuff you want to plot)
-aesthetics (aes() values- how the data map to the plot)
-geometries (geoms- how I want things drawn)
```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() +
  labs(title= "Speed and Stopping Distance of Cars", 
       x= "Speed (MPH)",
       y= "Stopping Distance (ft)",
       subtitle= "Ancient Data for Ancient Cars",
       caption= "Dataset: 'cars'") +
  geom_smooth(method="lm", se=FALSE) +
  theme_bw()
```


```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
```{r}
nrow(genes)
colnames(genes)
ncol(genes)
table(genes$State)
round(table(genes$State)/nrow(genes) * 100 , 2)
sum(genes$State == "up")
```

There are `r nrow(genes)` genes in this very serious dataset.
The names of the columns in this dataset are `r colnames(genes)` .
There are `r ncol(genes)` columns in this data set.
There are `r sum(genes$State == "up")` upregulated genes in this dataset.
```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2) +
  geom_point()
```
```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()



```
```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
p + scale_colour_manual(values=c("blue", "gray", "red")) 
```
```{r}
p + scale_colour_manual(values=c("blue", "gray", "red")) + labs(title= "Gene Expression Changes Upon Drug Treatment", 
       x= "Control (no drug)",
       y= "Drug Treatment")
```

## Section 7: gapminder

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
library("dplyr")
gapminder_2007 <- gapminder %>% filter(year==2007)
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) + geom_point(alpha=0.5)
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=pop) + geom_point(alpha=0.8)
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, size = pop) + geom_point(alpha=0.5)
```
```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x=gdpPercap, y= lifeExp,
                 size= pop), alpha= 0.5) + 
  scale_size_area(max_size = 10)
```

```{r}
library("dplyr")
gapminder_1957 <- gapminder %>% filter(year==1957)
ggplot(gapminder_1957) +
  geom_point(aes(x=gdpPercap, y=lifeExp, 
                 color = continent, size = pop), alpha=0.7) +
                   scale_size_area(max_size = 15)


                   
```
```{r}
gapminder_1957 <- gapminder%>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) +
  geom_point(aes(x= gdpPercap, y=lifeExp, color= continent, size = pop), alpha=0.7) +
  scale_size_area(max_size = 10) + 
  facet_wrap(~year)
```
```{r}
ggplot(mtcars) + 
  aes(x=mpg, y=disp) + 
  geom_point()
```

```{r}
ggplot(mtcars, aes(mpg, disp)) +
  geom_point()
```

