---
title: "Class 19: Pertussis Mini Project"
author: "Clarissa Savko (PID: A69028482)"
format: pdf
---

Pertussis is a bacterial infection that causes a severe cough known as "whooping cough". 

Let's have a look at case numbers of Pertussis in the US.

We can find the case numbers on the [CDC website] https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

```{r}
#| echo: FALSE
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
  
```

Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.
```{r}
library(tidyverse)
ggplot(cdc, aes(year, cases)) +
         geom_point() +
         geom_line()
```

Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes( x = year, y= cases) +
  geom_point() +
  geom_line()+
  geom_vline(xintercept = 1946, color= "red")+
  geom_vline(xintercept = 1996, color= "blue")+
  geom_vline(xintercept = 2019, color= "darkgreen")
```
Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There was an initial lag period for the first few years, than an increase in cases in the 2010's.

# CMI-PB Project

The CMI PB project collects and makes available data on the immune response to Pertussis booster vaccination. We will access this data via API. 
We will use the **jsonlite** package to access the data using the `read_json()` function.

```{r}
library(jsonlite)
subject <- read_json("https://cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject)
nrow(subject)
```
Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

60 aP and 58 wP subjects.

Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

There are 79 females and 39 males.

Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

There are several underrepresented groups including American Indians and Black.

```{r}
time_length(today() - mdy("11-26-1997"), "years")

```
Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?


Q8. Determine the age of all individuals at time of boost?

```{r}
subject$age <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
subject$age_years <- time_length(subject$age, "years")
subject$age_years
```


Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(age_years,
      fill=infancy_vac) +
  facet_wrap(vars(infancy_vac), ncol=1)+
  geom_histogram()
  
```

```{r}
specimen <- (read_json("https://www.cmi-pb.org/api/specimen", simplifyVector =TRUE))
head(specimen)

titer <- (read_json("https://www.cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE))
head(titer)
         
```
I want to merge (join) the specimen and subject tables together.

Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```
Now I want to merge the titer table with the meta table.

Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```
Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
igg <- abdata %>% filter(isotype == "IgG")
```

```{r}
ggplot(igg) + aes(MFI_normalised,
                  antigen) +
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
oops <- abdata %>% filter(antigen=="Fim2/3")
table(oops$dataset)
```

```{r}
table(abdata$dataset)
```

Select (or filter) for the 2021 dataset and isotype IgG

I want a time course of (`planned_day_relative_to_boost`) of IgG levels (`MFI_normalised`) for "PT" antigen. 

```{r}
igpt.21 <- abdata %>% filter (dataset== "2021_dataset", 
                   isotype == "IgG", 
                   antigen == "PT")
```

```{r}
ggplot(igpt.21) +
  aes(x = planned_day_relative_to_boost, 
      y = MFI_normalised,
      col = infancy_vac) +
  geom_point()+
  geom_line(aes(group = igpt.21$subject_id), linewidth=0.5, alpha=0.5) +
  geom_smooth(se=FALSE, span = 0.4, linewidth=3)
```



What about the 2022 dataset?


```{r}
igpt.22 <- abdata %>% filter (dataset== "2022_dataset", 
                   isotype == "IgG", 
                   antigen == "PT")
```

```{r}
ggplot(igpt.22) +
  aes(x = planned_day_relative_to_boost, 
      y = MFI_normalised,
      col = infancy_vac) +
  geom_point()+
  geom_line(aes(group = igpt.22$subject_id), linewidth=0.5, alpha=0.5) +
  geom_smooth(se=FALSE, span = 0.4, linewidth=3)
```

