---
title: "Class 10: Structural Bioinformatics (Pt. 1)"
author: "Clarissa Savko (PID:A69028482)"
format: pdf
---
```{r}
pdb_data <- read.csv("Data Export Summary.csv")
```
Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy

```{r}
pdb_data$Total <- as.numeric(sub("," ,"" , pdb_data$Total))
pdb_data$X.ray <- as.numeric(sub("," ,"" , pdb_data$X.ray))
pdb_data$EM <- as.numeric(sub("," ,"" , pdb_data$EM))
total_structures <- sum(pdb_data$Total)
total_X.ray <- sum(pdb_data$X.ray)
total_EM <- sum(pdb_data$EM)
(total_X.ray + total_EM) / total_structures * 100
```
93.16% of structures in the PDB are solved by X-ray and electron microscopy.

Q2: What proportion of structures in the PDB are protein?

```{r}
(pdb_data$Total[1]/total_structures) *100
```
86.67% of structures in the PDB are protein.

Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?

There are 7,434 HIV-1 protease structures in the current PDB.

Q4: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure?

Mol* is only showing the oxygen atom.

Q5: There is a critical “conserved” water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have

H20 308 seems to be the most involved in the binding site.

Q6: Generate and save a figure clearly showing the two distinct chains of HIV-protease along with the ligand. You might also consider showing the catalytic residues ASP 25 in each chain and the critical water (we recommend “Ball & Stick” for these side-chains). Add this figure to your Quarto document.

```{r}
library(imager)
molstar <- load.image("1HSG.png")
molstar
```

```{r}
library(bio3d)
pdb <- read.pdb("1hsg")
pdb
```
Q7: How many amino acid residues are there in this pdb object? 
There are 128 amino acid residues.
Q8: Name one of the two non-protein residues? 
Mk1 is a non-protein residue.
Q9: How many protein chains are in this structure? 
There are 2 protein chains in this structure.

```{r}
attributes(pdb)
head(pdb$atom)
```
```{r}
adk <- read.pdb("6s36")
adk
```
```{r}
m <- nma(adk)
plot(m)
```
```{r}
mktrj(m, file="adk_m7.pdb")
```


Q10. Which of the packages above is found only on BioConductor and not CRAN? 
"msa" is found only on Bioconductor and not CRAN.
Q11. Which of the above packages is not found on BioConductor or CRAN?: 
Grantlab/bio3d-view
Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?
True.
```{r}
library(bio3d)
aa <- get.seq("1ake_A")
aa
```
Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 

214 amino acids are in this sequence.

```{r}
#b <- blast.pdb(aa)
hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```
```{r}
#hits <- plot(b)
```
```{r}
head(hits$pdb.id)
```

```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
```{r}
ids <- basename.pdb(pdbs$id)
```

```{r}
#plot(pdbs, labels=ids)
```
```{r}
anno <- pdb.annotate(ids)
unique(anno$source)
anno
```
```{r}
pc.xray <- pca(pdbs)
plot(pc.xray)
```
```{r}
rd <- rmsd(pdbs)
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)
plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

```{r}
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

```{r}
library(ggplot2)
library(ggrepel)

df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```
```{r}
modes <- nma(pdbs)
plot(modes, pdbs, col=grps.rd)

```
Q14. What do you note about this plot? Are the black and colored lines similar or different? Where do you think they differ most and why?
They are different. They seem to differ the most in between residues 104 and 154 where the colored lines reach the highest peaks (fluctuations). I think this means those residues are the most flexible.